Enriching Object Detection with 2D-3D Registration and Continuous Viewpoint

### 2014-11-14 egpaper_final.tex
chris:cleaning up technical part

\section{Our Approach \scream{better name?}}
\label{sec:nz-who}

\subsection{Overview}
Our method attempts to match CAD models to 2D images, determining the best
model within a class, its pose and scale, and the focal length of the camera
used to produce the image. These parameters are first coarsely estimated using
an ensemble of templates generated offline, before they are refined using
MCMC by generating new detectors on the fly.

% The basic detection method underlying our approach is WHO template matching.
% However, we propose a new WHO variant called Non-Zero Whitened Histograms of
% Orientations (NZ-WHO), where the variant of WHO templates are made from 
% synthetic rendering to with

Our method In addition to the library of CAD models that will be matched to the query
images, our method requires a feature statistics which can be collected once and for all.

% images, our method requires a collection of random natural images. Offline, we
% compute HOG features of these images and record their mean and spatial
% autocovariance. The images are not used during testing, only the statistics
% generated from them.

Notably, we do not require any labelled training images of the objects to be
detected. Our templates are generated exclusively by rendering CAD models from
our library. From these renderings we produce novel HOG-like features we call
Non-Zero Whitened Histograms of Orientations (NZ-WHO), which use the natural
image statistics found offline to whiten a standard HOG template, making it
more discriminative.

The matching proceeds in two steps: First, we apply a bank of pre-trained
detectors to the image, corresponding to a set of discrete viewpoints, scales,
and object models. This provides a rough estimate of these parameters. Second,
we refine these parameters using MCMC, where each candidate is evaluated by
rendering a new whitened template on the fly.

Our ensemble of NZ-WHO templates can work as a detector but our algorithm 
can be initialized using proposal bounding boxes from
state-of-the-art object detectors \cite{Felzenszwalb10,Girshick14} before
executing the two steps above. In this scenario, our approach provides a means
to augment standard bounding box detectors with more interesting 3D
information.
\begin{comment}
Our method is in line with \cite{Aubry14, Lim14} in the sense that we
use only renderings, not real images. However, rather than using fixed
sized mid-level patches, we use whole objects and make root
templates. By using root templates, there is no need to learn a
discriminative patch dictionary or to calibrate the patches and an
additional inference stage is not required. However, whitening the
whole template is prohibitive for several reasons. All templates have
different aspect ratios making it difficult to generate the covariance
matrix $\Sigma$. Since the root templates are larger than mid-level
patches, a very large auto-covariance $\Gamma$ is required. Lastly,
time complexity of decorrelating a template with $n$ cells increases
as $O(n^3)$, resulting in slow template generation and large residuals
due to numerical errors accumulated while decomposing the covariance
matrix $\Sigma$.

To overcome these difficulties, we propose the Non-Zero Whitened
Histograms of Orientations (NZ-WHO) with iterative Conjugate Gradient to
generate templates in real time and enrich rectangular boxes with high
quality 2D-3D matching. The 2D-3D matching provides a CAD model,
sub-category, rendering, depth, viewpoint and focal length.

Our method can work as a standalone 2D-3D matching algorithm, or even
as a standalone ensemble of exemplar-based object detectors (Figure
\ref{fig:front}) but its real power comes from combining generic
object detectors and complement rectangular bounding boxes. Our method
is simple in design; is several orders of magnitude faster than
traditional method; requires only CAD models; does not need neither images nor
calibration, yet is powerful in giving high-quality metadata. These
properties make our method ideal for complementing object detections.

To complement detection bounding boxes, we first search possible 2D-3D
matching by running cached discrete viewpoint templates. Then, for the
most probable match, we refine the initial matching using Metropolis
Hastings. Since it requires iterations of proposal and validation, we
present real time analysis of NZ-WHO template generation to show that
this can be done on-the-fly. The fine tuning stage process jointly
optimizes translation, yaw, roll, pitch, scale, and focal length
continuously and also searches for different CAD models.

\end{comment}
\subsection{Rendering Pipeline}

@@

\subsection{Non-Zero Whitened Histograms of Orientations (NZ-WHO)}

Recently, Hariharan \etal. introduced Whitened Histograms of Orientations (WHO)
\cite{Hariharan12}, which uses statistics from natural images to whiten HOG
templates, making them more discriminative. Whitening is a common signal
processing operation for decorrelating a set of random variables
\cite{Martinsson05, Belouchrani00}. More formally, suppose that we have a
$k$-dimensional random variable $X \in \mathcal{R}^k$ with $Cov(X)=\Sigma$. By
whitening the signal,
\begin{equation}
\tilde{X}=\Sigma^{-\frac{1}{2}}(X - E[X]) \label{eq:whitening}
\end{equation}
, we
remove 2nd order correlation between all features. So, all elements in the
random variable $\tilde{X} = \left[\tilde{X}_1, \tilde{X}_2, \dots,
\tilde{X}_k\right]$, become more independent. In the context of HOG features, a
whitened HOG template will produce Gaussian noise when convolved with a random
natural image, but a delta-like response on an image matching the template. WHO
templates are more discriminative than standard HOG templates because they only
capture the gradients that differentiate the object from random images.

Hariharan \etal. explained the
whitening as Linear Discriminant Analysis (LDA) where all classes have different
means but have the same covariance.

@@

We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), that are both
more discriminative than WHO, and can be generated several orders of magnitude
faster (in 70ms compared to several seconds for WHO). This speed up means we
can generate templates on the fly, allowing our approach to evaluate arbitrary
viewpoint hypotheses.

In \cite{Hariharan12}, the authors compute the mean and covariance matrix by assuming Wide-Sense Stationarity (WSS) of HOG features
generated from natural images. That is, they assume the mean of a HOG cell is
independent of its place in the image, and the autocovariance of cells depends
only on their relative location.

In one dimension, let $x(u)$ be the feature at location $u$. Then WSS states
that $\mathbb{E}\left[x(u)\right] = \mu$ for all $u$, and
\begin{equation}
\textrm{cov}_x(u,v) = \textrm{cov}_x(0, v-u) = \Gamma(v-u),
\end{equation} where $\Gamma$ is the autocovariance. For simplicity, we describe the 1D case but this can be easily extended to 2D spatial autocovariance.

Therefore, assuming WSS allows the covariance matrix of templates \emph{of any
size} to be synthesized from the autocovariance matrix using a simple lookup.

Our method requires generating a covariance matrix of non-zero cells

\subsubsection{Whitening Synthesized Templates}
Our first innovation is `Non-Zero' whitening. When synthesizing detection templates from rendered images, a common problem is how to handle the background. If the model is rendered over a natural image background, gradients in the background will be incorporated into the discriminative template, potentially causing false-positives if background elements exist in the query image.

Alternatively, if the background is left textureless \scream{did Hariharan do this?} (see Fig.~\ref{fig:rendering}), whitening the resulting HOG template will produce undesirable background artifacts. These are created when centering the template (by subtracting the mean $\mu$), where strong negative weights are introduced in the textureless region (as seen in Fig.~\ref{fig:whocomparison}). This could result in positive matches being suppressed due to spurious background gradients.

\scream{might need to match notation with subsequent sections}
NZ-WHO removes these artifacts so that the background has no effect on the template response. Given HOG features $x$, we create a new vector $x'$ which contains only the non-zero elements of $x$. We select the same elements of $\mu$ to produce $\mu'$, while the autocovariance lookup is only performed for the cells with non-zero gradient, producing $\Sigma'$ corresponding to the features in $x'$. After solving the resulting system (which is now smaller than in the WHO approach) we find
\begin{equation}
w'=\Sigma'^{-1}(x' - \mu') \label{eq:nz-who},
\end{equation}which we reassemble into the final NZ-WHO template $w$ by replacing the non-zero elements of $x$ with the corresponding $w'$ element.

Notice that $\Sigma$ is not square-rooted in Eq.~\ref{eq:nz-who}, whereas it is in the definition of whitening (Eq.~\ref{eq:whitening}). This is because the whitened template will only be used in the context of detection, which involves the inner product with the whitened HOG features of the query image $y$:
\begin{equation}
(y-\mu)^{T}\Sigma^{-\frac{1}{2}}\Sigma^{-\frac{1}{2}}(x-\mu)\\
= (y-\mu)^{T} w.
\end{equation}
We can therefore capture the whitening of the query image features in $w$.

@@

%\scream{chris:We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), an improvement over WHO which is both more discriminative than WHO and can be computed many orders of magnitude faster than WHO (70ms as opposed to to several seconds for WHO). This speedup allows us to generate NZ-WHO templates on the fly, allowing us to quickly evaluate arbitrary viewpoint hypotheses.}

@@

The covariance matrix for HOG templates is typically ill-conditioned, but we can add a regularization constant to the diagonal to reduce its condition number to the point where the conjugate gradient method will converge quickly. We use a constant of $0.15$, which reduces the condition number from $10^{20}$ to $50$, much smaller than the dimension of the matrix (7000).

As a result, a GPU implementation of conjugate gradient converges in \scream{x
ms}, two orders of magnitude faster than the system can be solved using the
Cholesky factorization (\scream{5 s}).

@@

% \subsection{Non-Zero Whitened Histograms of Orientations (NZ-WHO)}
% why non zero whitening and why it makes more sense than seeing 3D, IKEA
% Overview of the subsections
%   Generating covariance quickly
%   Fast inversion
%   Regularization

% Overview of the subsections
% In this section, we will give a brief overview of whole whitening stages. In the first section \ref{feature_statistics}, we will discuss about first and second order statistics of HOG feature. Then, we will discuss about using autocovariance $\Gamma$ to create covariance matrix $\Sigma$ for non-zero cells and SIMD implementation to speed up the generation. Then, we will explain fast and accurate way to solve $w = \Sigma^{-1}x$ using Conjugate Gradient method. % Finally, we will analyze the effect of regularization.

@@

\begin{figure}[t]
  \begin{center}
     \includegraphics[width=0.9\linewidth]{covariancetime} 
  \end{center}
  \caption{Time to generate $\Sigma$ for various number of cells (left) and same plot on log scale(right). The time complexity increases as $O(n^2)$ for $n$ number of cells.}
  \label{fig:covariancetime_crop}
\end{figure}

@@

\begin{figure}[t]
  \begin{center}
  \begin{tabular}{cc}
     \includegraphics[width=0.5\linewidth]{whotime} & 
     \includegraphics[width=0.5\linewidth]{speedup}\\
     (a) & (b) \\
 \end{tabular}
  \end{center}
  \caption{Comparison of HOG variants generation time and final }
  \label{fig:whotime}
\end{figure}
[Figure spectral analysis]

>>>>

\section{Approach Overview}
\label{sec:nz-who}
\input{overview}

\subsection{Rendering}

We used a public rendering engine to create realistic rendering and
depth\begin{comment}\cite{Choy14render}\end{comment}. The CAD models we used
in our experiments contain color, texture and material information such as
transparency and reflectance. We tried to make the rendering as much realistic
as possible to simulate the natural image statistics. The example rendering is
in Fig. \ref{fig:rendering}. We also extracted depth information as well while
rendering to use in the reconstruction.

@@

\begin{figure}[t]
  \begin{center}
     \includegraphics[width=0.4\linewidth]{rendering}
     \includegraphics[width=0.4\linewidth]{depth}
  \end{center}
  \caption{Example rendering and depth made by our renderer}
  \label{fig:rendering}
\end{figure}

\subsection{Whitened Histograms of Orientations (WHO)}

Recently, Hariharan \etal. introduced Whitened Histograms of Orientations (WHO)
\cite{Hariharan12}, which uses statistics from natural images to whiten HOG
templates, making them more discriminative. Whitening is a common signal
processing operation for decorrelating a set of random variables
\cite{Martinsson05, Belouchrani00}. More formally, suppose that we have a
$k$-dimensional random variable $X \in \mathcal{R}^k$ with $Cov(X)=\Sigma$. By
whitening the signal,
\begin{equation}
\tilde{X}=\Sigma^{-\frac{1}{2}}(X - E[X]) \label{eq:whitening}
\end{equation}
, we remove 2nd order correlation between all features. Since covarinace
collection is expensive, \cite{Hariharan12} proposed easy way to generate
covariance $\Sigma$ from autocovariance $\Gamma$

@@

% In \cite{Hariharan12}, the authors compute the mean and covariance matrix by assuming Wide-Sense Stationarity (WSS) of HOG features
% generated from natural images. That is, they assume the mean of a HOG cell is
% independent of its place in the image, and the autocovariance of cells depends
% only on their relative location.

% In one dimension, let $x(u)$ be the feature at location $u$. Then WSS states
% that $\mathbb{E}\left[x(u)\right] = \mu$ for all $u$, and
% \begin{equation}
% \textrm{cov}_x(u,v) = \textrm{cov}_x(0, v-u) = \Gamma(v-u),
% \end{equation} where $\Gamma$ is the autocovariance. For simplicity, we describe the 1D case but this can be easily extended to 2D spatial autocovariance.
% 
% Therefore, assuming WSS allows the covariance matrix of templates \emph{of any
% size} to be synthesized from the autocovariance matrix using a simple lookup.

% However, the 
% We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), that are both
% more discriminative than WHO, and can be generated several orders of magnitude
% faster (in 70ms compared to several seconds for WHO). This speed up means we
% can generate templates on the fly, allowing our approach to evaluate arbitrary
% viewpoint hypotheses.

% Our method requires generating a covariance matrix of non-zero cells

\subsubsection{Whitening Synthesized Templates and Non-Zero WHO}
Our first innovation is `Non-Zero' whitening. When synthesizing detection
templates from rendered images, a common problem is how to handle the
background. If the model is rendered over a natural image background, gradients
in the background will be incorporated into the discriminative template,
potentially causing false-positives if background elements exist in the query
image.

Alternatively, if the background is left textureless (see
Fig.~\ref{fig:rendering}), whitening the resulting HOG template
will produce undesirable background artifacts. These are created when centering
the template (by subtracting the mean $\mu$), where strong negative weights are
introduced in the textureless region (as seen in Fig.~\ref{fig:whocomparison}).
This could result in positive matches being suppressed due to spurious
background gradients.

NZ-WHO removes these artifacts so that the background has no effect on the
template response. Given HOG features $x$, we create a new vector $x'$ which
contains only the non-zero elements of $x$. We select the same elements of
$\mu$ to produce $\mu'$, while the autocovariance lookup is only performed for
the cells with non-zero gradient, producing $\Sigma'$ corresponding to the
features in $x'$. After solving the resulting system (which is now smaller than
in the WHO approach) we find

\begin{equation}
w'=\Sigma'^{-1}(x' - \mu') \label{eq:nz-who},
\end{equation}which we reassemble into the final NZ-WHO template $w$ by
replacing the non-zero elements of $x$ with the corresponding $w'$ element.

% Notice that $\Sigma$ is not square-rooted in Eq.~\ref{eq:nz-who}, whereas it is
% in the definition of whitening (Eq.~\ref{eq:whitening}). This is because the
% whitened template will only be used in the context of detection, which involves
% the inner product with the whitened HOG features of the query image $y$:
% \begin{equation}
% (y-\mu)^{T}\Sigma^{-\frac{1}{2}}\Sigma^{-\frac{1}{2}}(x-\mu)\\
% = (y-\mu)^{T} w.
% \end{equation}
% We can therefore capture the whitening of the query image features in $w$.

@@

%\scream{chris:We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), an improvement over WHO which is both more discriminative than WHO and can be computed many orders of magnitude faster than WHO (70ms as opposed to to several seconds for WHO). This speedup allows us to generate NZ-WHO templates on the fly, allowing us to quickly evaluate arbitrary viewpoint hypotheses.}
\scream{add comments about the fig:whocomparison}

@@

The covariance matrix for HOG templates is typically ill-conditioned\cite{Hariharan12}, but we
can add a regularization constant to the diagonal to reduce its condition
number to the point where the conjugate gradient method will converge quickly.
We use a constant of $0.15$, which reduces the condition number from $10^{20}$
to $50$, much smaller than the dimension of the matrix (7000).

As a result, a GPU implementation of conjugate gradient converges in 50
ms when using 250 HOG cells, two orders of magnitude faster than the system can be solved using the
Cholesky factorization.

\begin{figure}[t]
  \begin{center}
  \begin{tabular}{cc}
     \includegraphics[width=0.5\linewidth]{whotime} & 
     \includegraphics[width=0.5\linewidth]{speedup}\\
     (a) & (b) \\
 \end{tabular}
  \end{center}
  \caption{(a) Comparison of WHO variants generation time and HOG extraction time (b) final speed of using NZ-WHO compare to WHO}
  \label{fig:whotime}
\end{figure}

We compare the speed of each of template generation methods in
Fig.~\cite{fig:covariancetime_crop} Using naive Cholesky decomposition,
WHO template takes several seconds. But, if we use iterative Conjugate Gradient method, it only
takes 100ms. If we use NZ-WHO, we can gain extra speed up since we only whitens non-zero cells.

@@

% \subsection{Non-Zero Whitened Histograms of Orientations (NZ-WHO)}
% why non zero whitening and why it makes more sense than seeing 3D, IKEA
% Overview of the subsections
%   Generating covariance quickly
%   Fast inversion
%   Regularization
% Overview of the subsections
% In this section, we will give a brief overview of whole whitening stages. In the first section \ref{feature_statistics}, we will discuss about first and second order statistics of HOG feature. Then, we will discuss about using autocovariance $\Gamma$ to create covariance matrix $\Sigma$ for non-zero cells and SIMD implementation to speed up the generation. Then, we will explain fast and accurate way to solve $w = \Sigma^{-1}x$ using Conjugate Gradient method. % Finally, we will analyze the effect of regularization.

@@

% \begin{figure}[t]
%   \begin{center}
%      \includegraphics[width=0.9\linewidth]{covariancetime} 
%   \end{center}
%   \caption{Time to generate $\Sigma$ for various number of cells (left) and same plot on log scale(right). The time complexity increases as $O(n^2)$ for $n$ number of cells.}
%   \label{fig:covariancetime_crop}
% \end{figure}

@@

[Figure spectral analysis]
